Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-10-09 text/manuscript-full-KS.qmd
added cerpp proposal

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758_blinded; @cottom2017lower; @RN4324].

@@

```{=html}
<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->
```
This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board Enrollment Planning Service (EPS), which advises colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```
The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. The [Online Appendix](https://structuring-college-access.github.io/) describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family income greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->
```
Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggest that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^8] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

@@

Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.
--> 
```
**RQ2**. Our second research question asks, how does the racial and socioeconomic composition of included versus excluded prospects vary when student list purchases filter on Geomarkets? We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from these search criteria.

@@

-->
```
# Results

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```
@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 23% of all prospects but 51% of Black students. Similarly, South and Southwest Suburbs accounts for 17% of all prospects but 26% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 77% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 11% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the more than 2,500 Hispanic students (45% versus 23% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

@@

Analyses of student list orders across the Dallas-Fort Worth metropolitan area reveals patterns of racial/ethnic and socioeconomic stratification in Geomarket contributions. As Dallas-Fort Worth became increasingly diverse from 1980 to 2020, a more balanced racial/ethnic distribution emerged across the metropolitan area's Geomarket landscape. However, these patterns did not translate to racial/ethnic composition of student prospect pools. Suburban Geomarkets like Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex yielded a disproportionately larger share of Asian prospects, whereas central Geomarkets such as the City of Dallas, Dallas County, and Irving, Arlington, and Grand Prairie contributed a disproportionately larger share of Hispanic and Black prospects. These racial/ethnic disparities become significantly more pronounced, particularly in the disproportionate share of Asian prospects, as PSAT score thresholds increase (see Appendix @fig-dallas-rq2-lowSAT-race and @fig-dallas-rq2-highSAT-race). Similar Geomarket contribution patterns are also evident by first-generation college student status, with Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex providing larger shares of non-first generation college student prospects, and central Geomarkets (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie) providing disproportionately larger shares of first-generation college student prospects. Moreover, central Geomarkets are particularly important for including Black (City of Dallas and City of Fort Worth), Hispanic (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie), and Asian (City of Fort Worth and Irving, Arlington, and Grand Prairie) first-generation college students. While Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex Geomarkets still contribute a notable overall number of Black, Hispanic, and Asian prospects, such students are predominantly not first-generation college students.
-->
```
## Los Angeles

@@

 In 1980, poverty rates were relatively low (less than 12%) for all Geomarkets except for East Los Angeles (17%), South and South Central Los Angeles (24%), and Hollywood and Wilshire (18%). However, poverty rates declined in South and South Central Los Angeles (22%), remained stable in East Los Angeles and Hollywood and Wilshire, and increased in all other Geomarkets. Such shifts and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in more staggered poverty rates across the metropolitan area by 2020. This includes four Geomarkets with poverty rates less than 10% (San Fernando Valley- West, West Los Angeles and West Beach, South Bay, and Covina and West Covina), four Geomarkets with poverty rates ranging from 11%-15% (San Fernando Valley- East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario), and three Geomarkets with poverty rates ranging from 16%-22% (Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles). -->
```
<!--# Lastly, @fig-la-rq1-ses shows a widening disparity in educational attainment across the Los Angeles metropolitan area. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the West Los Angeles and West Beach Geomarket experienced a substantially greater increase - from 35% in 1980 to 66% in 2020- widening the gap with other Geomarkets in terms of educational attainment. -->

@@

```{=tex}
\begingroup
\small
```
::: {#refs}
:::

```{=tex}
\endgroup
\clearpage
\singlespacing
```
<!-- TABLES -->

@@

```{=tex}
\clearpage
```

>>>>

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```

Sociology offers several perspectives to describe the process of sorting students into colleges as a process that tends toward social reproduction [@RN4873; @RN2377]. On the student demand side, the status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary to claim spots at selective colleges [e.g., @RN1228; @RN1241; @RN4991]. <!--@RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. -->On the supply-side, the credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required to claim advantageous social and economic positions [@RN1297], and this race for credentials benefits affluent households [e.g., @RN865; @RN861]. Scholarship on enrollment management offers an agentic depiction of college behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758_blinded; @cottom2017lower; @RN4324].

@@

```{=html}
<!--  We make third-party EM products the object of empirical analysis, though we cannot claim to assess how the population of colleges utilize these products.
# REMOVED FROM PARAGRAPH ABOVE: First, in the case of college rankings, extant research shows how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges that are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375].  -->
```

This manuscript analyzes the Market Segment Model [@RN4982], which categorized high school students into vertical market segments -- local, in-state, regional, national -- and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board Enrollment Planning Service (EPS), which advises colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables by @RN5005, we cannot show the effect of quantification on social reproduction. However, by demonstrating which students are included versus excluded in student list purchases that filter on particular Geomarkets, our analyses provide novel insight into the underlying mechanism by which quantification can reproduce historical class-based and race-based inequality in the sorting of students into colleges.

@@

```{=html}
<!-- SAT test-takers were categorized into one of four different market segments -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. 
A "regional" student submits more SAT scores to regional institutions than they do to local, in-state, or national institutions. A "national" student submits more SAT scores to national institutions than they do to local, in-state, or regional institutions.
-->
```

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. The [Online Appendix](https://structuring-college-access.github.io/) describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family income greater than \$35,000, percent with both parents having a BA, etc.[^5] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^6] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

@@

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->
```

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggest that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^8] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

@@

Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.
--> 
```

**RQ2**. Our second research question asks, how does the racial and socioeconomic composition of included versus excluded prospects vary when student list purchases filter on Geomarkets? We answer this question using actual student lists purchased by public universities. In February 2020, we began issuing public records requests to public universities about student lists purchased from 2016 through 2020. We narrowed the scope of our request to student lists purchased from College Board, ACT, and the National Research Center for College and University Admissions (NRCCUA), the three largest student list vendors at the time. For each student list purchased from 2016 through 2020, we requested two related pieces of data: (1) the order summary, which specifies search criteria for the student list purchase; and (2) the de-identified prospect-level list produced from these search criteria.

@@

-->
```

# Results

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```

@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 23% of all prospects but 51% of Black students. Similarly, South and Southwest Suburbs accounts for 17% of all prospects but 26% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 77% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 11% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the more than 2,500 Hispanic students (45% versus 23% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

@@

Analyses of student list orders across the Dallas-Fort Worth metropolitan area reveals patterns of racial/ethnic and socioeconomic stratification in Geomarket contributions. As Dallas-Fort Worth became increasingly diverse from 1980 to 2020, a more balanced racial/ethnic distribution emerged across the metropolitan area's Geomarket landscape. However, these patterns did not translate to racial/ethnic composition of student prospect pools. Suburban Geomarkets like Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex yielded a disproportionately larger share of Asian prospects, whereas central Geomarkets such as the City of Dallas, Dallas County, and Irving, Arlington, and Grand Prairie contributed a disproportionately larger share of Hispanic and Black prospects. These racial/ethnic disparities become significantly more pronounced, particularly in the disproportionate share of Asian prospects, as PSAT score thresholds increase (see Appendix @fig-dallas-rq2-lowSAT-race and @fig-dallas-rq2-highSAT-race). Similar Geomarket contribution patterns are also evident by first-generation college student status, with Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex providing larger shares of non-first generation college student prospects, and central Geomarkets (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie) providing disproportionately larger shares of first-generation college student prospects. Moreover, central Geomarkets are particularly important for including Black (City of Dallas and City of Fort Worth), Hispanic (City of Dallas, City of Fort Worth, Dallas County, and Irving, Arlington, and Grand Prairie), and Asian (City of Fort Worth and Irving, Arlington, and Grand Prairie) first-generation college students. While Collin and Rockwall Counties and the West of Dallas/Fort Worth Metroplex Geomarkets still contribute a notable overall number of Black, Hispanic, and Asian prospects, such students are predominantly not first-generation college students.
-->
```

## Los Angeles

@@

 In 1980, poverty rates were relatively low (less than 12%) for all Geomarkets except for East Los Angeles (17%), South and South Central Los Angeles (24%), and Hollywood and Wilshire (18%). However, poverty rates declined in South and South Central Los Angeles (22%), remained stable in East Los Angeles and Hollywood and Wilshire, and increased in all other Geomarkets. Such shifts and West of Dallas/Ft. Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in more staggered poverty rates across the metropolitan area by 2020. This includes four Geomarkets with poverty rates less than 10% (San Fernando Valley- West, West Los Angeles and West Beach, South Bay, and Covina and West Covina), four Geomarkets with poverty rates ranging from 11%-15% (San Fernando Valley- East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario), and three Geomarkets with poverty rates ranging from 16%-22% (Hollywood and Wilshire, East Los Angeles, South and South Central Los Angeles). -->
```

<!--# Lastly, @fig-la-rq1-ses shows a widening disparity in educational attainment across the Los Angeles metropolitan area. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the West Los Angeles and West Beach Geomarket experienced a substantially greater increase - from 35% in 1980 to 66% in 2020- widening the gap with other Geomarkets in terms of educational attainment. -->

@@

\begingroup
\small

::: {#refs}
:::

\endgroup
\clearpage
\singlespacing

<!-- TABLES -->

@@

\clearpage
